---
title: "Introduction to R Part 1"
author: "Arjun Singh Yadav"
date: "28 April 2017"
output: html_document
---

# Introduction to R part 1

### Key Features of R:
Open Source and Free

Powerfull and Scalable

Lightweigt and Multi-platform

### Extensibility of R:
Can connect to databases as Oracle, MySQL

Can call C, C++ and Fortran code

Can be used as an embeded computing engine (Rserver)

Can be used to deploy Interactive apps (Shiny)

### Drawbacks:
Data is read in Memory 

R is single threaded, limiting speed and efficiency

## What would we learn in this session
#### R Objects
Function call and variables

Primitive data types

Vectors Matrices and arrays

Data Frames

Lists 

#### Fuctional Programming:apply


```
#Installing any package in R can be done by

#install.packages("packagename")

#importing the package in the environment can be done by

#library(packagename)

```
### Variables

Variable is a symbolic name if a storage location

```{r , echo=TRUE}
x=2
print(x)

```


### Function Call

As all other languages like python, java,sql, function takes a argument, given
a parentheses and return values, multiple arguments can be seperated by commas.

Some functions are built into R. When we call a function, just as when you
write any other expression, RStudio prints result.

```{r , echo=TRUE}
sin(0)
log(100,10)
x <- pi/2 
print (x)
```


### Primitive Data Types

Numeric: The data type for real number, A number is assumed by R to be of type
"numeric" even if it is an integer.

Character: The data type for a string.

Logical: TRUE OR FALSE

Factor: Categorical variable that cna be either numeric or string variable.
These repressent groups.

Integer: 7L where 7 is a integer, 7 is a real number while 7.0 is a float type.

```{r , echo=TRUE}
#to check data type use
class(2)
```


### Logical Operators
AND
```
1==1 & 2==3
```

OR
```{r , echo=TRUE}
1==1 | 2==3
```

NOT
```{r , echo=TRUE}
!(1==1)
```


### Type Converstion
```{r , echo=TRUE}
as.integer(5.6)
as.character(5.6)
```

There is one more way using
```{r , echo=TRUE}
#as.type()
```


### Data Objects

Vectors: Sequece of primitive data types.
```{r , echo=TRUE}
c(1,2,3,4)
c('ad','cd','ef')
c('ab',3)

x <- c(1,2,3)
y <- c('ab','x')
c(x,y)

seq(1,10)

seq(1,10,by=2)

seq(1,10,length=5)

rnorm(10)

runif(5)

sample(c('A','B'), size=10, replace=TRUE)

rep(0,10)

rep(c(1,2,3),5)

rep(c(1,2,3),c(3,5,1))

#slicing
x = c(1,2,3,4,5)
x[1]

#naming of vector
v = 10:15
names(v) = c('a', 'b', 'c', 'd', 'e', 'f')
v

v['a']

#vector computation
c(1,2,3,4) + c(5,6,7,8)

c(1,2,3,4) + c(1,2) 

#Logical computation
x = c(1,2,3,4)
x[ x>=2 & x<4 ] 

x = c(1,2,3,4)
exp(x)

max(x)
min(x)
mean(x)

```

#### Exercise
Find your current working directory

Get documentation for function 'lm'

Run function ls(), check what does it print out?

Check how to remove everythign in a workspace by using rm().

Assign the text "Hello world by" to x

Assign a y variable with your name

Use function cat(), learn what it does, try cat(x,y)


### Matrices

#### Matrix: Two Dimentional collection of primitve data objects.

```{r , echo=TRUE}
my_mat = matrix(1:12,4,3)
print (my_mat)


my_mat = matrix(1:12, nrow=4, ncol=3)

v <- 1:12
dim(v) <- c(4, 3) 
print (v)

#cbind()
vector1 <- 1:4
vector2 <- 5:8
vector3 <- 9:12
cbind(vector1, vector2, vector3)

#Matrix slicing
my_mat
#try
my_mat[1,2]

my_mat[1, 1:3] # first row, all columns

my_mat[-1, ] # omit first row; all columns

my_mat[1, ]

my_mat[ , 2]

my_mat[ , 2, drop=F] #if you want to keep the structure in mat

#operations on matrices
my_mat + 1

(my_mat + 1) - my_mat

mean(my_mat)

```

### Arrays

Array: Multi- dimentional collection of premitive data objects.
```{r , echo=TRUE}
a <- array(1:24, dim = c(3,4,2))
print(a)
```

### DataFrames

Data frames are spreadsheets like datastructure in which data type of different
columns can be different.

Both columns and rows are labeled, columns always have descriptive names and
rows are labeled by sequential indices. 


DataFrame: Matrix with labelled columns and rows. Like a spreadsheet

#### Creating Dataframe
```{r , echo=TRUE}
city = c('Seattle', 'Chicago', 'Boston', 'Houston')
temp = c(78, 74, 50, 104)

citydf <- data.frame(city=city, temp=temp)
citydf

citydf <- data.frame(city, temp)
citydf

colnames(citydf) = c('CITY', 'TEMP')
citydf

citydf <- data.frame(city, temp,
 row.names=c('a', 'b', 'c', 'd'))
citydf

citydf = data.frame(city, temp)

citydf[3,]

citydf[, 1, drop=F]

#my fav type 

citydf$city

```

#### Selecting and Slicing Data Frames

```{r , echo=TRUE}
ave = mean(citydf$temp)

citydf[citydf$temp > ave, ]

```

#### Function on DataFrames
```{r , echo=TRUE}
dim(citydf)
head(citydf)
str(citydf)
summary(citydf)
order(citydf$temp) #order returns row indices in increasing.

citydf[order(citydf$temp),]

```

#### Sorting DataFrames
```{r , echo=TRUE}
order(citydf$temp, decreasing=TRUE)

citydf[order(citydf$temp, decreasing=TRUE),]
```

#### Exporting and Importing a DataFrame
```{r , echo=TRUE}
#Exporting
write.csv(citydf, file='my_data.csv', row.names=F)

#Importing
read.csv('my_data.csv')

```

### Lists

List: Collection of names values, possibly of different types

List have heterogenous data, it is also the most flexible type of R object.

```{r , echo=TRUE}
#Creating list
employee = list(Name='John', Position='cashier',Salary=9.50)
print (employee)


student <- list(Name="Peter",Classes =c("Bio101","Psych200","CS125"))
 
print(student)

#slicing 
student[2]

class(student[2])

student['Name'] # or student$Name

employee[-1] 

employee[c("Name", "Position")]

employee$Name

employee[[1]]

#Modifying a list

employee$Position = 'manager'
employee[[3]] = 10.50
employee


```

### Functional Programming

```{r , echo=TRUE}
data(iris)

v = 1:4
sapply(v, sqrt)

sapply(iris[ ,1:4], mean) #return a vector

lapply(v, mean) #returns the value in list

#apply
mat = matrix(1:6, nrow=2, ncol=3)

apply(mat, 1, sum) # across rows

apply(mat, 2, sum) # across columns

```
################################ END #######################################


# Introduction to R part 2

### More Data Types
Charaters

Factors

### Control Statement
Conditional

Loops

### Functions

### Data Transformations 

#### Charater Manupulation

➢ nchar(): Count the number of characters. 

➢ strsplit(): Split the elements of a character vector.

➢ paste(): Concatenate strings. 

➢ substr(): Create substrings of a character vector.

➢ gsub(): Replace parts of strings. 

➢ grep(): Complete pattern matching.

#### Character Functions: nchar
```{r , echo=TRUE}
fruit = 'apple orange grape banana'
nchar(fruit)

test = '10+$ 0f $ymb01$!'
nchar(test)
```

#### Character Functions: strsplit

```{r , echo=TRUE}
fruit = 'apple orange grape banana'
strsplit(fruit, split=' ')

fruitvec = strsplit(fruit, split=' ')[[1]]
print(fruitvec)
```

#### Character Functions: paste

```{r , echo=TRUE}
paste(fruitvec, c('A', 'B'))

paste(fruitvec, c('A', 'B', 'C'), sep='::')

paste(fruitvec, collapse=',')

paste(fruitvec, c('A', 'B'), sep=':', collapse='@')
```

#### Character Functions: gsub

```{r , echo=TRUE}
gsub('apple', 'strawberry', fruit)

gsub('a', '?', fruit)

gsub('an', 'HA', fruit)

```
#### Character Functions: grep

```{r , echo=TRUE}
grep('grape', fruitvec)
grep('a', fruitvec)
grep('an', fruitvec)

```
### Factors 
Factors represent categorical values or discrete values that might be used as discriminators in statistical test.

```{r , echo=TRUE}
data(iris)
species = iris$Species 
class(species)

print(species)

levels(species)

class(levels(species))

ranks <- c(4, 1, 1, 4, 3, 3, 2, 3, 2, 4)
rank_factor = factor(ranks)
rank_factor

# CUT
#The cut() function is used to divide an integer vector into intervals to be
#used as factors.  cut() divides a numeric vector into a specified number of
#interval sections:

x = c(1, 2, 3, 4, 5, 6, 7, 8) 
f = cut(x, 3)
class(f)

print(f)

```

### Control Statement
#### Conditionals

```{r , echo=TRUE}
num = 5
if (num %% 2 != 0) {
  cat(num, 'is odd') 
}

num = 4
if (num %% 2 != 0) {
  cat(num, 'is odd')
  } else {
  cat(num, 'is even')
}

# more loops as For , if else , while etc

```

### Functions

```{r , echo=TRUE}
name = function(arguments) {
  body
  return(expression)
}
```

### Data Transformations

#### Merge and Split
 “Data transformation” is a very general term that could be used to reference
 many operations on data. In data science and engineering, it generally refers
 to non-numerical operations that rearrange and combine data.
 
We will see many data transformations in R. In a future class, we will look at
dplyr, an entire package devoted to transformations.
 
 Here we introduce a few widely used functions for combining data frames and
 splitting them apart.


#### Merge Two Data Frames

```{r , echo=TRUE}

genders = data.frame(id=c(1,2,3), gender=c('M', 'F', 'M'))
names = data.frame(id=c(3,1,2), name=c('tom','john','mary'))

merge(genders, names, by='id')
```

#### Split into Groups
```{r , echo=TRUE}
iris_split = split(iris, iris$Species) 
class(iris_split)

attributes(iris_split)

str(iris_split)

##### Use function Unsplit to undo what split performed.
iris_unsplit <- unsplit(iris_split, iris$Species)
class(iris_unsplit)

iris_unsplit[c(1, 51, 101), ]
```

###################################### END ###################################

# Dplyr
Dplyr packageprovides functions for effective manipulation of data frames
1) Filtering and rearranging.
2) Joins
3)Groupwise operations- aggregation

### Basic dplyr functions
Filter - take subset of rows

Select - Take subset of columns

Arrange - Reorder Rows

Mutate/ transmute - Add or replace columns

Summarise - Compute aggregate values

Various Joins - Inner, outer, Full , Semi

#### Baby birth dataset
```{r}
births <- read.csv("data/births.csv",stringsAsFactors = FALSE)
births[c(1:2, 131:132),]

bnames <- read.csv("data/bnames.csv",stringsAsFactors = FALSE)
head(bnames, 5)

mike = bnames[bnames$name == "Mike", ] 
head(mike, 3)
 
```
### Built in Functions
#### Filter and Select

```{r, echo=TRUE}
df = data.frame(color = c("blue","black", "blue", "blue","black"),value = 1:5)

print(df)

#import the package
library(dplyr)

# Filter
filter(df, color == "blue")

#you can also achieve the same result by doing
df[df$color == "blue",]

filter(df, value %in% c(3, 4, 5)) # note no $ sign is needed to call column value

filter(df, value %in% c(3, 4, 5) & color=='blue')

filter(df, value %in% c(1, 4) | color=="black")

# Select
select(df, color)

select(df, color, value)

select(df, -color)

# Select can also be used with coditions
#starts_with(x, ignore.case=T)
#ends_with(x, ignore.case=T)
#contains(x, ignore.case=T)
#matches(x, ignore.case=T)
#one_of(name_1, name_2, name_n)

#Example
select(df, starts_with('c'))

# Select and Rename

df = rename(df, COLOR = color)

```
#### Arrange and Multate
 Arrange positions the rows of the dataframe in ascentding order according to
 values in specific columns
 
```{r, echo=TRUE}
df = data.frame(color = c("blue","black", "blue", "blue","black"),value = 1:5)
arrange(df, value)
 
arrange(df, desc(value)) 
 
arrange(df, color)
 
# Mutate
#Mutate fuction creates a new column by derivinf its value from old column
 
mutate(df, double =value*2)
 
mutate(df, double = 2 * value, quadruple = 4 * value)

#Transmute
# Transmute only keeps the new column and deletes the old ones

transmute(df, double = 2 * value, quadruple = 4 * value)

```
### Summarise

We often need to extract summary information about columns, for finding mean or
sd of feature.
```{r, echo=TRUE}
summarise(df, total = sum(value)) 

summarise(df, total = sum(value),avg = mean(value))

# you can use min() max() mean() sum() sd() median()
# n():returns the number of observations
#n_distinct(x): returns the number of unique values in the column
#first(x) last(x): returns the first or last observations in the column x
#nth(x,n): returns the nth observation in the column x

summarise(df, n_distinct(color))

summarise(df, nth(value, 4))

```

### Join data sets

```{r, echo=TRUE}
x = data.frame( name = c("John", "Paul", "George",
                          "Ringo", "Stuart", "Pete"),
                instrument = c("guitar", "bass", "guitar",
                                "drums", "bass", "drums"),
                stringsAsFactors = FALSE)
print(x) 

y <- data.frame( name = c("John", "Paul", "George",
                          "Ringo", "Brian"),
                band = c(TRUE, TRUE, TRUE,
                         TRUE, FALSE),
                stringsAsFactors = FALSE)
print(y)

#Inner Join
inner_join(x, y, by = "name")

# Left Join

left_join(x, y, by = "name")

right_join(x, y, by = "name")

full_join(x, y, by = "name")

semi_join(x, y, by = "name")


```
### Group By and Summarise

####Group_by

```{r , echo=TRUE}
by_color = group_by(df, color) 
class(by_color) 

summarise(by_color, total = sum(value))

```

### The %>% operator

```{r , echo=TRUE}
# group by and summarise
library(dplyr)

bnames <- read.csv("data/bnames.csv",stringsAsFactors = FALSE)

group_by(bnames, name, year)

bnames %>% group_by(., name, year)

bnames %>% group_by(name, year)

```

############################ END ###########################################

# GGPLOT2

```{r iris, echo=TRUE}
# a simple plot without help of ggplot2
data(iris)

plot(iris$Sepal.Width, iris$Sepal.Length)

```

# Grammar of Graphics
data: R dataframe
aesthetics: values that will be plotted, eg. x and y values, plus color and
shape of dots for a scatter plot.

geoms: the type of plot

facets: side by side plots

stats: statistical transformation like binning and smoothing

scales: adjust sizes

coordinates: what coordinate system to use

themes: overall presentation

### Constructing a ggplot2 plot

```{r , echo=TRUE}
library(ggplot2)

#Scatter plot

#mpg
head(mpg)

g <- ggplot(data = mpg, aes(x = displ, y = hwy))
g + geom_point()

#AESTHETICS 
g + geom_point(aes(color = class))
g + geom_point(aes(size = class))
g + geom_point(aes(shape = class))
g + geom_point(aes(alpha = class))

# plotting one of these
g + geom_point(aes(color = class))

# Faceting
#facet_grid(): 2d grid, rows ~ cols, . is a placeholder that can be used if you
#              don't want a split.

#facet_wrap(): 1d ribbon wrapped into 2d.

g + geom_point() + facet_grid(. ~ cyl)
g + geom_point() + facet_grid(drv ~ .)
g + geom_point() + facet_grid(drv ~ cyl)
g + geom_point() + facet_wrap( ~ class)

#Geoms
# they create statistical plotting over scatter plots, showing a trend
g + geom_smooth() # method = "auto" is the default

# this case plots both the stats and the points
g + geom_point() + geom_smooth()

#you can use different methods in stats
g + geom_point() + geom_smooth(method = "lm")

#example

ggplot(data = mpg, aes(x = class, y = hwy)) + geom_point()


g <- ggplot(data = mpg, aes(x = class, y = hwy))
g + geom_boxplot()

# BOXPLOTS

ggplot(data = mpg, aes(x = reorder(class, hwy), y = hwy)) + geom_boxplot()

ggplot(data = mpg,aes(x=reorder(class,hwy,median), y = hwy)) + geom_boxplot()

#There is a lot we can do in ggplots refer to  http://docs.ggplot2.org/

```

########################## END ###############################################



